# DATA PREPROCESSING

In this step, we are setting our **R** system by importing and loading libraries require in this project. Once all important libraries will be installed, we are importing and understanding our rainfall input data.

## Checking  and loading libraries

Here we are checking **R** libraries and if not found in system then we install it.Libraries which are getting used are as below:

 1. **Data Manipulation**: `data.table`,`dplyr`,`sqldf` and `reshape`
 2. **Descriptive Statistics**: `pastecs`,`psych`,`tidyquant`,`nortest` and `tseries`
 3. **Data Visulization**: `ggthemes`,`ggplot2` and `scales`
 4. **Forecasting**: `forecast` and `Metrics`
 5. **Reporting**: `kableExtra` and `pander`

```{r setup}
list.of.packages <-c("data.table","dplyr","plyr","sqldf",
                     "reshape","pastecs","psych","tidyquant",
                     "nortest","tseries","ggthemes","ggplot2",
                     "scales","forecast","Metrics","kableExtra",
                     "pander","tidyverse","lubridate","timeSeries",
                     "magrittr","recipes")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

if(length(new.packages)) install.packages(new.packages)
pander::pander("All libraries are updated and installed")

```


## Load libraries 

Here we load libraries.

```{r message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)
library(plyr)
library(sqldf)
library(reshape)
library(pastecs)
library(psych)
library(tidyquant)
library(nortest)
library(tseries)
library(ggthemes)
library(ggplot2)
library(scales)
library(forecast)
library(Metrics)
library(kableExtra)
library(pander)
library(tidyverse)
library(lubridate)
library(timeSeries)
library(magrittr)
library(recipes)

pander("All libraries are successfully installed")
```

## Data Preprocessing

In this step, we are importing data and understanding it.We have obtained this data by secondary research from this source [Indian Institute of Tropical Meteorology](https://tropmet.res.in/static_pages.php?page_id=53).

In this step, we are importing input data and exploring it by some descripitive statistics and graphs.  


### *Data Import*

Let's start by loading our data set.For importing the data, we are using `read.csv` function from the base **R** packege.
```{r message=FALSE, warning=FALSE}
setwd("D:/A Notes by 1st sem MSc(Nitesh)/Project/project 2/Final")
rainfall=read.csv('rainfall.csv')

pander("Data loaded successfully")

```

### *Structure of data*

In this step, we are checking structure of our **rainfall** dataset.

```{r message=FALSE, warning=FALSE}
# Stucture of data
str(rainfall)
```

**Findings:**

 > 1. Dataframe consists of 146 observation and 18 variables.
 
 > 2. All variables are numeric except YEAR variable which is integer.

